Haskell in Computational Biology

### 2013-05-27 everygen.tex
paper changes

everyPt :: RandomGen r
        => SearchGen pt r -> CCost -> Scored pt
        -> Rand r [CCosted (Utility (Scored pt))]
everyPt sg cost startPt = do
  successors <- mapM (nextPt sg) (repeat startPt)
  tagged <- zipWithM costedUtility successors [succ cost..]
  let (useless, CCosted (Useful newPt) newCost : _) =
                      span (isUseless . unCCosted) tagged
  (++) (CCosted (Useful startPt) cost : useless) <$>
                                  everyPt sg newCost newPt
 where costedUtility pt cost =
         utility sg move >>= \u -> return $ CCosted u cost
        where move = Move { older = startPt, younger = pt
                          , youngerCCost = cost }

>>>>

everyPt :: RandomGen r
        => SearchGen pt r -> CCosted (Scored pt)
        -> Rand r [Utility (CCosted (Scored pt))]
everyPt sg (CCosted cost startPt) = do
  successors <- zipWithM costedUtility [succ cost..] =<<
                mapM (nextPt sg) (repeat startPt)
  let (useless, Useful pt : _) = span isUseless successors
  (++) (Useful (CCosted cost startPt) : useless) <$> everyPt sg pt
 where
   costedUtility cost pt = return . fmap (CCosted cost) =<< utility sg move
     where move = Move { older = startPt, younger = pt, youngerCCost = cost }

### 2013-05-27 jfp.tex
paper changes

@@

A~model is represented as a sequence of \emph{nodes}, 
each of which includes three states:
\smallverbatiminput{model3-node}
The representation of a state~$s$ includes the scores of transitions leaving
that state;
if~state~$s$ is a match or insertion state, its representation also
includes a table of emission scores.
For~example, the representation of a match state is \texttt{MState}:
\smallverbatiminput{model3-mstate}
%
The transition scores, which are shown as $a_{s\hat s}$
in \figref{plan7}, are read by  
function \texttt{aScore}, 
%which is given states $s$~and~$\hat s$ and a node index~$j$; its
whose
specification is
\mbox{$\mathtt{aScore}\;s\;\hat s\;j = \txprobxj s {\hat s}$}.
The
emission table is read by 
\texttt{eScore}, whose specification
is % parallel structure for line breaking below
\mbox{$\mathtt{eScore}\;s\;j\;i = e'_{s_j}(x_i)$},
where $x_i$ is the amino acid located at position~$i$ in the query
sequence. 
%
%%  We place the transition probabilities 
%%  into a record in which each field is labeled
%%  \texttt{$s$\_$\mskip 1mu\hat s$},
%%  where $s$~and~$\hat s$ form one of the 7~permissible pairs of state
%%  labels:\label{sec:tprobs}
%%  \smallverbatiminput{tprob-tprobs}
%%  \remark{Need to explain those last two fields. HMMer?}

@@

Another of us was curious to try coding Viterbi's equations
directly as recursive functions.
Like a recursive Fibonacci function, Viterbi's functions,
if~implemented \naive ly,
take exponential time.
But like the Fibonacci function, Viterbi's functions can be
\emph{memoized}.
For example, to compute $V_j^{\prime M}(i)$ using the equation at the top
of \figref{viterbi-transformed}, we define
\mbox{\texttt{vee\char`\'} \texttt{Mat} $j$ $i$}.
The equation adds~$e^{\prime}_{M_{j}}(x_{i})$, computed
with \texttt{eScore}, to a minimum of sums.
The~sum of an~$a'_{s\hat s}$~term and a $V^{\prime s}_{j-1}(i-1)$ term 
is computed by function \texttt{avSum}, in which
the terms are computed by \texttt{aScore} and \texttt{vee''}, respectively:
\smallfuzzverbatiminput{2.9pt}{vfix}
%%%%%%%%%% horrible line break
% The right-hand side of \verb+vee'+ is wrapped in a call to
What about the call to
\mono{\mbox{fmap (Mat `cons`)}}?
This call performs a computation \emph{not} shown
in \figref{viterbi-transformed}:
\mrfy\ computes not only the {probability} of the most likely path but 
also the path itself.
Function \mono{(Mat `cons`)} adds~$M$ to a path;
we~avoid \mbox{\mono{(Mat :)}} for reasons explained
in \secref{performance} below.

@@

If~\mrfy\ could apply Hughes's technique, it~would search by composing
a \emph{generator} 
that produces an infinite sequence of moves, 
a \emph{filter} that selects the useful moves, and 
a \emph{test} function that enumerates finitely many useful moves
and returns the final destination.
But a generator may produce an infinite sequence of useless moves.
(For example, if \mrfy\ should stumble upon a global optimum, every move from
that point would be useless.)
Given an infinite sequence of useless inputs, a filter would not
produce any values, and the search would diverge.
This situation is depicted graphically in
\figref{filtered-search-moves}:
on the left-hand side, 
the first move is useless; the second move is useful; and all
subsequent moves are useless.
Filtering out useless moves would produce the sequence shown on the
right-hand side, which contains two points followed by a divergent
computation.

@@

To tell if a point returned by \texttt{nextPt} is useful, we call
 the \texttt{utility} function,
which scrutinizes a move represented as follows:
\smallverbatiminput{move}
The decision about utility uses not only a source of randomness but
also the \emph{cumulative cost} of the point, which we define to be
the number of points explored previously.
The~cumulative cost of the current point is also the age of the
search,
and
in~simulated annealing, for example, as the search ages,
the \texttt{utility} function becomes less likely to accept a move
that worsens the score.

\noindent
\vbox{
Using these pieces, function \texttt{everyPt} produces an infinite
sequence containing a mix of useful and useless moves:
\smallfuzzverbatiminput{10.8pt}{everygen}
}
\noindent
Both \texttt{nextPt} and \texttt{utility} are monadic, but 
we can still exploit laziness:
from its starting point, \texttt{everyPt} produces an infinite list of
randomly chosen successor points, then calls \texttt{costedUtility} to
tag each one with a cumulative cost and a utility.

In Hughes's \citeyearpar{hughes:why} classic model of lazy search,
we~might expect to use an infinite sequence of points obtained by
applying \texttt{nextPt} to many points in succession, as done by
the Prelude function \texttt{iterate}.
But~if you look carefully at how \texttt{successors} is computed,
that's not what's happening:
instead, we~compute an infinite list by applying \texttt{nextPt} to
a~\emph{single} initial point using infinitely many distinct sources
of random variation.
\remark{This is the place to bring in a figure, I~think. ---NR}
Then, when we find a useful \texttt{newPt}, a~recursive call
to \texttt{everyPt} computes a \emph{fresh} infinite sequence of
single jumps starting at~\texttt{newPt}.
\mrfy's search may explore an arbitrary number of infinite
lists of points, and the representation we~found easiest to think
about is the pair of elements
\texttt{pt0}~and~\texttt{nextPt} in the \texttt{SearchGen} record.

Once the successors have been computed and tagged,
\texttt{span} finds the first useful successor.
In~case there \emph{is} no successor, \texttt{everyPt} also returns
all the useless successors.
If~we do find a useful successor, we start searching anew from that
point, with a recursive call to \texttt{everyPt}.
(Because \texttt{everyPt} is monadic, the points accumulated so far are
appended to its result using the \verb+<$>+~operator.)
%
The most informative part of \texttt{everyPt} is last expression of
the \texttt{do} block,
which shows that the result begins with a useful point,
is followed by a (possibly infinite, possibly empty) list of useless
points, and then continues recursively with another call to \texttt{everyPt}.

%%\begingroup\parskip=0.9\parskip
%% \makeatletter\@sectionaboveskip=0.9\@sectionaboveskip\makeatother
The rest of the search uses Hughes's
classic composition of generator  and test function.
Because our code is monadic, we~use the monadic composition
operator~\texttt{=<<}, which is the bind operator with its arguments
swapped: 
\smallverbatiminput{search}
The \texttt{test} function has type \mono{SearchStop~pt}:
\smallverbatiminput{stop}
Type \mono{History~pt} retains only the useful points.
(Internally, \mrfy\ needs only the \emph{final} useful point,
but because we want to study how different search algorithms behave,
we~keep all the useful points.)

@@

We also benchmarked \mrfy\ using a model of an ``8-bladed beta propeller.''
The model has 343~nodes, of which 178 appear in 40~beta strands.
The segments between beta strands typically have at most 10~nodes.
We~used a query sequence of 592 amino acids, but each placement breaks
the sequence into 41~pieces, each of which typically has at most 20 amino
acids.
Because \mrfy\ can solve the models between the beta strands independently,
this benchmark has a lot of parallelism,
which Haskell made it easy to exploit.
Using
\texttt{Control.Parallel}, parallelizing the computation was as easy
as  substituting
\texttt{parmap rseq} for \texttt{map}.
\figref{fig:speedup} shows speedups when using from 
 1~to~48 of the cores 
on a 48-core, 2.3GHz AMD Opteron 6176 system.
Errors are estimated from 5~runs.
After about 12~cores, where \mrfy\ runs 6~times as fast as sequential
code, speedup rolls off.
By~running 4~instances of \mrfy\ in parallel on different searches,
we~hope to be able to use all 48~cores with about 50\%~efficiency.

@@

@@

\subsection{Competing with C++; less FORTRAN; more Haskell}

EXPLAIN SECTION TITLE.

WHY IS THIS SECTION RELEVANT?  TO SHOW TRAPS AND PITFALLS FOR NEW
HASKELL PROGRAMMERS.

FIRST STEP WAS TO SIMPLIFY THE CODE.

As shown in \figrefpair{viterbi-original}{viterbi-transformed} and in
the code for \verb+vee'+ on page~\pageref{code:vee-prime}, Viterbi's
$V'$ is indexed by a residue index~$i$, a~node index~$j$,
and
a state $M$, $I$, or~$D$.
A~query sequence is simply an array of~\texttt{AA} indexed by~$i$,
and a model is an array of nodes indexed by~$j$.
The node array and its indices are handled as in our group's C++ code:
all nodes have the same representation, and every node can be
interpreted in one of the special states Begin and End.
(This possible interpretation causes the introduction of the
mysterious probabilities \verb+b_m+ and \verb+m_e+ into the transition
table on page~\pageref{sec:tprobs}.)
These conventions of representation lead to a full implementation of
Viterbi's algorithm that is more reminiscent of FORTRAN than of
Haskell:
\begin{itemize}
\item
Recursive computations are done on smaller array indices, not smaller
models.
\item
Array indices 1, 0, and $-1$ are all used to indicate special
conditions.
\item
Senseless combinations of array indices and states are handled by
assigning each such combination the largest representable score, which
corresponds to the smallest representable probability.
\end{itemize}
This is the code in which \mrfy\ spends 98\%\ of its time, and that
we~wanted to speed up to make it more competitive with~C++.
But the central function \verb+vee'+ has 25~cases overall.
Before thinking seriously about how to tune it, we~decided to simplify~it.

Our first step in simplification was to start thinking like Haskell
programmers and make more effective use of both the static type system:
(AND THE PATTERN-MATCH COMPILER)
\begin{itemize}
\item
We introduced a \texttt{Model} type that consists of a single Begin
node, an~unbounded sequence of Middle nodes, and a single End node.
\item
We~observed that the main algorithm where \mrfy\ spends all its time
uses only Match, Insert, and Delete states.
We~redefined the \texttt{StateLabel} type to include only these
states; the Begin and End labels are now part of
type \texttt{FullStateLabel}. 
\end{itemize}
As a result of these changes, in the revised code,
\begin{itemize}
\item
Only array index~0 is used to indicate a special
conditions: the transition from the leftmost middle node to the unique
begin node.
\item
All well-typed combinations of array indices and states make sense;
senseless combinations are ruled out by the static type system.
\end{itemize}

\begin{figure}
\def\vsum#1#2{#1+#2}
\[
V_{j}^{\prime \hat s}(i) = e^{\prime}_{\hat s_{j}}(x_{i}) 
  + \min\limits_s \left\{
       \vsum{V_{\mathit{pnode}\;\hat s\;j}^{\prime
       s}(\mathit{presidue}\;\hat s\;i)} {a^{\prime}_{s \,\hat
       s}}, \mbox{ where $s$ can precede $\hat s$} \right \}
\]

\caption{Viterbi's transformed equations, abstracted over states $s$
and $\hat s$, defining $e^{\prime}_D(x) = 0$}
\figlabel{viterbi-abstracted}
\end{figure}

We also combined Viterbi's three equations into one.
Each equation has the same form, but is parameterize over three
values determined by the right-hand state~$\hat s$:
the~index of the previous node ($j$~or~$j-1$),
the~index of the previous residue ($i$~or~$i-1$),
and the set of previous states~$s$ over which the minimum is taken.
If~we abstract the index computations into functions
$\mathit{pnode}$ 
and 
$\mathit{presidue}$
respectively,
we~get the combined, abstract form of Viterbi's equations shown in
\figref{viterbi-abstracted}.
The~implementation of \figref{viterbi-abstracted} looks like this:
\smallverbatiminput{hoviterbi}
Function \texttt{next} makes the recursive call to

The right-hand sideOur 
Indices 1, 0, and $-1$ were all used to indicate special conditions,
and

\subsubsection{Reducing cases in Viterbi}

\newcommand\pv{\textit{Predecessor Viterbi}}
\newcommand\ov{{Original Viterbi}}

Our first step in adapting the code was to create \pv.
Like \ov, \pv\ takes a state and two indices as an argument.
But at top level, there is no case analysis on the state, and the only
indices considered are zero and nonzero, so there are only four cases.
(The state is used in two ways: to compute transition probabilities,
and to decide which states may permissibly precede the current state.
The set of permissible preceding states is looked up in a data
structure, not computed by case analysis.)

Here is the general case in the new code:

THERE IS A GRATUITOUS CHANGE: THE TYPE OF ARRAY USED IS DIFFERENT.
WE SHOULD PROBABLY FIX THAT.
In the old code, there are three corresponding cases: one each for
match, insert, and delete states.  
Here is the code for the match case:
\smallverbatiminput{viterbi}
The other case are similar; the second argument to \texttt{aScore} is
always equal to the first (state) argument to \texttt{vee'},
and the argument to \texttt{map from} is different in each case.
(In the new code, this argument is identical to \texttt{preceders !
stateRight}.)
NORMAN THINKS THIS COMPARISON DOESN'T WORK AT ALL :-(

The main benefit is that the Viterbi function is reduced from 90~lines
of code to just 40~lines.
This change laid the groundwork for more experiments.

@@

\section{Notes on the new paper}

>>>>

\newcommand\vfilbreak[1]{\vskip 0pt plus #1 \penalty-200 \vskip 0pt plus -#1 }

@@

A~model is represented as a sequence of \emph{nodes}, 
each of which includes three states:\label{code:model3-node}
\smallverbatiminput{model3-node}
The representation of a state~$s$ includes the scores of transitions leaving
that state;
if~state~$s$ is a match or insertion state, its representation also
includes a table of emission scores.
For~example, the representation of a match state is \texttt{MState}:
\smallverbatiminput{model3-mstate}
%
The transition scores, which are shown as $a_{s\hat s}$
in \figref{plan7}, are read by  
function \texttt{aScore}, 
%which is given states $s$~and~$\hat s$ and a node index~$j$; its
whose
specification is
\mbox{$\mathtt{aScore}\;s\;\hat s\;j = \txprobxj s {\hat s}$}.
The
emission table is read by 
\texttt{eScore}, whose specification
is % parallel structure for line breaking below
\mbox{$\mathtt{eScore}\;s\;j\;i = e'_{s_j}(x_i)$},
where $x_i$ is the amino acid located at position~$i$ in the query
sequence. 
%
%%  We place the transition probabilities 
%%  into a record in which each field is labeled
%%  \texttt{$s$\_$\mskip 1mu\hat s$},
%%  where $s$~and~$\hat s$ form one of the 7~permissible pairs of state
%%  labels:\label{sec:tprobs}
%%  \smallverbatiminput{tprob-tprobs}
%%  \remark{Need to explain those last two fields. HMMer?}

@@

Another of us was curious to try coding Viterbi's equations
directly as recursive functions.
Like a recursive Fibonacci function, Viterbi's functions,
if~implemented \naive ly,
take exponential time.
But like the Fibonacci function, Viterbi's functions can be
\emph{memoized}.
For example, to compute $V_j^{\prime M}(i)$ using the equation at the top
of \figref{viterbi-transformed}, we define
\mbox{\texttt{vee\char`\'} \texttt{Mat} $j$ $i$}.
The equation adds~$e^{\prime}_{M_{j}}(x_{i})$, computed
with \texttt{eScore}, to a minimum of sums.
The~sum of an~$a'_{s\hat s}$~term and a $V^{\prime s}_{j-1}(i-1)$ term 
is computed by function \texttt{avSum}, in which
the terms are computed by \texttt{aScore} and \texttt{vee''},
respectively:\label{code:vee-prime} 
\smallfuzzverbatiminput{2.9pt}{vfix}
%%%%%%%%%% horrible line break
% The right-hand side of \verb+vee'+ is wrapped in a call to
What about the call to
\mono{\mbox{fmap (Mat `cons`)}}?
This call performs a computation \emph{not} shown
in \figref{viterbi-transformed}:
\mrfy\ computes not only the {probability} of the most likely path but 
also the path itself.
Function \mono{(Mat `cons`)} adds~$M$ to a path;
we~avoid \mbox{\mono{(Mat :)}} for reasons explained
in \secref{performance} below.

@@

If~\mrfy\ could apply Hughes's technique, it~would search by composing
a~\emph{generator} 
that produces an infinite sequence of moves, 
a~\emph{filter} that selects the useful moves, and 
a~\emph{test} function that enumerates finitely many useful moves
and returns the final destination.
But~a~generator may produce an infinite sequence of useless moves.
(For example, if \mrfy\ should stumble upon a global optimum, every move from
that point would be useless.)
Given an infinite sequence of useless inputs, a~filter would not
produce any values, and the search would diverge.
This situation is depicted graphically in
\figref{filtered-search-moves}:
on the left-hand side, 
the first move is useless; the second move is useful; and all
subsequent moves are useless.
Filtering out useless moves would produce the sequence shown on the
right-hand side, which contains two points followed by a divergent
computation.

@@

To tell if a point returned by \texttt{nextPt} is useful, we call
 the \texttt{utility} function,
which scrutinizes a move represented as follows:
\smallverbatiminput{move}
The decision about utility uses not only a source of randomness but
also the \emph{cumulative cost} of the point, which we define to be
the number of points explored previously.
The~cumulative cost of the current point is also the age of the
search,
and
in~simulated annealing, for example, as the cumulative cost increases,
the \texttt{utility} function becomes less likely to accept a move
that worsens the score.
To~attach a cumulative cost to a point, we use type \texttt{CCosted}:
\smallverbatiminput{aged}

Using these pieces, function \texttt{everyPt} produces an infinite
sequence containing a mix of useful and useless points.
Each useful point is tagged with cumulated cost and its score:
\smallfuzzverbatiminput{6pt}{everygen}
Both \texttt{nextPt} and \texttt{utility} are monadic, but 
we can still exploit laziness:
from its starting point, \texttt{everyPt} produces an infinite list of
randomly chosen successor points, then calls \texttt{costedUtility} to
tag each one with a cumulative cost and a utility.
Because we want to compose monadic computations in the same order we
compose ordinary functions, 
our code uses the monadic composition
operator~\texttt{=<<}, which is the bind operator with its arguments
swapped.

If \mrfy\ used Hughes's classic model of lazy search,
we~could possibly expect to see an infinite sequence of points obtained by
applying \texttt{nextPt} to many points in succession, as~done by
the Prelude function \texttt{iterate}.
But~that's not how \mrfy\ works;
instead, 
the~infinite list \texttt{successors}
is computed by applying \texttt{nextPt} to
the \emph{single} initial point \texttt{startPt}, \mbox{using} infinitely
many distinct sources of random variation.
To~understand the structure of
 \mrfy's more complex search, it~may help to
relate the code to \figref{search-moves}.
Please assume that \texttt{startPt} is the point in the upper left
corner of \figref{search-moves}.
Then the first part of the \texttt{successors} list appears in the
next column of \figref{search-moves}:
the~first four elements are \texttt{useless};
the next element, which is the first useful point, is~\texttt{pt};
and~the remaining elements, which are never evaluated,
match the
underscore.
The first useful point is located by \mono{span~isUseless}.
After the search reaches~\texttt{pt},
we~start the search anew with a recursive 
call to \mono{everypt~sg}, 
passing~\texttt{pt}.
%
The most informative part of \texttt{everyPt} is last expression of
the \texttt{do} block,
which shows that the result begins with a useful point,
is followed by a (possibly infinite, possibly empty) list of useless
points, and then continues with the results of a recursive call
to \texttt{everyPt}.

\mrfy's search may explore an arbitrary number of infinite
lists of points, each reached using a different source of randomness.
We~suppose this search space could be represented by a single
value of type \mbox{\mono{Rand r [[Scored pt]]}}, but when writing
client code, we~found it much easier to think about the
representation we actually use in the
\texttt{SearchGen} record:
the~monadic value
\texttt{pt0} together with the function \texttt{nextPt}.

%%\begingroup\parskip=0.9\parskip
%% \makeatletter\@sectionaboveskip=0.9\@sectionaboveskip\makeatother

\vfilbreak{2.5\baselineskip}

The rest of the search uses Hughes's
classic composition of generator  and test function.
\smallverbatiminput{search}
The \texttt{test} function has type \mono{SearchStop~pt}:
\smallverbatiminput{stop}
Type \mono{History~pt} retains only the useful points.
(Internally, \mrfy\ needs only the \emph{final} useful point,
but because we want to study how different search algorithms behave,
we~keep all the useful points.)

@@

We also benchmarked \mrfy\ using a model of an ``8-bladed beta propeller.''
The model has 343~nodes, of which 178 appear in 40~beta strands.
The segments between beta strands typically have at most 10~nodes.
We~used a query sequence of 592 amino acids, but each placement breaks
the sequence into 41~pieces, each of which typically has at most 20 amino
acids.
Because \mrfy\ can solve the models between the beta strands independently,
this benchmark has a lot of parallelism,
which Haskell made it easy to exploit.
Using
\texttt{Control.Parallel}, parallelizing the computation was as easy
as  substituting
\texttt{parmap rseq} for \texttt{map}.
\figref{speedup} shows speedups when using from 
 1~to~48 of the cores 
on a 48-core, 2.3GHz AMD Opteron 6176 system.
Errors are estimated from 5~runs.
After about 12~cores, where \mrfy\ runs 6~times as fast as sequential
code, speedup rolls off.
By~running 4~instances of \mrfy\ in parallel on different searches,
we~hope to be able to use all 48~cores with about 50\%~efficiency.

@@

\subsection{Improving \mrfy's sequential performance}

Research software for
computational biology doesn't have to be fast to be useful.
But~when we say that \mrfy\ solves some problems 8~times slower then
SMURF, we~felt it wasn't a good showing for Haskell.
\mrfy\ spends 98\%\ of its time running Viterbi's algorithm, and
we~felt sure that with some love and attention, we could make it
more competitive with~C++.

The code example for \verb+vee'+ on page~\pageref{code:vee-prime}
shows the computation of the general case of~$V^{\prime M}_j(i)$ from 
\figref{viterbi-transformed}.
But the full story is not nearly as nice at one could hope for:
in~the code that accompanies our 2012 ICFP paper,
\verb+vee'+ has 25~cases overall.
Many of these cases exist because we~did not have enough insight into
Haskell, and we made some decisions that might have been more
appropriate to a FORTRAN program:
\begin{itemize}
\item
Recursive computations were done on smaller array indices, not smaller
models.
\item
Array indices 1, 0, and $-1$ were all used to indicate special
conditions.
\item
Senseless combinations of array indices and states were handled by
assigning each such combination the largest representable score, which
corresponds to the smallest representable probability.
\end{itemize}
Before trying to speed up the code, we~simplified~it.

Our first step in simplification was to start thinking like Haskell
programmers and use types:
\begin{itemize}
\item
We~refactored the single, flat \texttt{Node} type from our ICFP paper
into the three state types you see on page~\pageref{code:model3-node}.
\item 
We introduced a \texttt{Model} type that consists of a special Begin
node containing a begin and an insertion state, an~unbounded sequence
of ordinary \texttt{Nodes}, and an end state.
\item
We~observed that the main algorithm where \mrfy\ spends all its time
uses only match, insertion, and deletion states, so we removed the
value constructors corresponding to begin and end states.
Those constructors were used only to prettyprint models, so~we made
them part of a larger type.
\end{itemize}
As a result of these changes, in the revised code,
\begin{itemize}
\item
Only array index~0 is used to indicate a special
condition: the transition from the leftmost middle node to the unique
begin node.
\item
All well-typed combinations of array indices and states make sense;
senseless combinations are ruled out by the static type system.
\end{itemize}

\def\p#1{\mathit{p#1}}
\begin{figure}
\def\vsum#1#2{#1+#2}
\[
\begin{array}{r@{}c@{}l}
\multicolumn3{c}{
  V_{j}^{\prime \hat s}(i) = e^{\prime}_{\hat s_{j}}(x_{i}) 
    + \min\limits_s \big\{
         \vsum{V_{\mathit{\p j}}^{\prime s}(\p i\; s)} 
     {a^{\prime}_{s_{\p j} \,\hat s_j}}, 
      \mbox{ where $s$ can precede $\hat s$}  \big\}
}
\\[10pt]
\mbox{where }\p i\; s&{} = {}& 
  \left\{ \begin{array}{l}
          i \mbox{, if $s=D$}\\
          i-1 \mbox{, otherwise}\\
          \end{array}
  \right.
\\[10pt]
\p j &{} = {}& 
  \left\{ \begin{array}{l}
          j \mbox{, if $\hat s=I$}\\
          j-1 \mbox{, otherwise}\\
          \end{array}
  \right.
\\
\end{array}
\]

\caption{Viterbi's equations, abstracted over state~$\hat s$, 
defining $e^{\prime}_{D_j}(x) = 0$}
\figlabel{viterbi-abstracted}
\end{figure}

Our next step was to use more abstraction.
In~Viterbi's equations,
$V'$ is indexed by a residue index~$i$,\footnote
{In~biology, when an amino acid enters a protein, it often loses a few
atoms, and so within the protein, the amino acid is often called
a \emph{residue}.}
 a~node index~$j$,
and
a state $M$, $I$, or~$D$.
Each state uses a slightly different equation, but it is possible to
combine them.
\figref{viterbi-abstracted} shows a single equation which, when
state $\hat s$ is specialized to state $M$, $I$, or~$D$, recovers the
original three equations from \figref{viterbi-transformed}.
The figure sums the cost of emitting an amino acid at state~$\hat s$
lowest of the costs of the paths entering~$\hat s$ from predecessor
state~$s$.
Index~$\p j$ is the index of the node containing the predecessor
state, and $\p i; s$ is the index of the amino acid, if~any, to be
emitted at state~$s$.

NEED TO EXPLAIN PRECEDERS

\figref{viterbi-abstracted} doesn't deal with the begin state.
In~our legacy file format, and also in our code, the begin state is
represented as a match state inside an ordinary node.
In~a perfect world, then, we would add one case to say that the match
state with $i=j=0$ is reachable at zero cost, and we'd be finished.
But the world is imperfect, because although the begin state is
represented as a match state, unlike a match state, the begin state
does not emit.
We~therefore need a special case for each of the three states that can
follow the begin state.
Here is the code:
\smallverbatiminput{hov4}
NEED TO EXPLAIN INTERNAL, EDGE, AND LEAF.
NEED TO EXPLAIN [[transition]] and [[emission]]

\smallverbatiminput{hov-prevs}

The main benefit is that the Viterbi function is reduced from 90~lines
of code to just 38~lines, of which 8~lines are type signatures.
This change laid the groundwork for more experiments.

If~we look not at the whole package but just at Viterbi's algorithm,
the~performance story is even more grim:
we~wrote a custom C~program that computes a score using Viterbi's
algorithm and does nothing else, and it runs about a hundred times
faster than \mrfy's implementation.

THE STORY OF THE UNPACK PRAGMAS; CITING JOHAN TIBELL

@@

@@

\section{THIS SECTION WILL BE DELETED.  SAUVE QUI PEUT!}
